The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-09-02 thesis/introduction.tex
julio sep 1 edits to §1, 2

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These galaxies are typically defined to be
fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
or similarly \(M_\star \lesssim 10^9 M_\odot\)
\citep[e.g.,][]{hodge1971, mcconnachie2012}. Because the galaxy
luminosity function increases towards fainter objects, dwarfs are the
most numerous galaxies in the Universe
\citep[e.g.,][]{blanton+2005, mao+2021}. Dwarf galaxies are also highly
\emph{dark-matter dominated}, with mass to light ratios which may exceed
1000 \(M_\odot/ L_\odot\) \citep[e.g.,][]{simon+geha2007, hayashi+2023}.
With the exception of the Magellanic Clouds, most dwarf galaxy
satellites of the Milky Way (MW) are \emph{quenched}, with little to no
recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
satellites contain stellar populations which are \emph{relics} from the
early universe, consisting of many of the oldest and most metal poor
stars. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
small scales to the formation of metal-poor stellar populations.

In this section, we first describe the general observed properties of
local dwarf galaxies. Next, we summarize the cosmological origin and
role of dwarf galaxies. We later review recent advancements and problems
concerning dwarf galaxies. Next, we contextualize the extended stellar
density profiles of Sculptor and Ursa Minor. We then discuss how tides
influence satellite galaxies in general. Finally, we provide a roadmap
to the thesis.

\section{Observational history of dwarf
galaxies}\label{observational-history-of-dwarf-galaxies}

Dwarf galaxies have long raised conundrums for theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) are also classified as dwarf galaxies, but these
  were likely always known to humans at southern latitudes.}, with no
known analogues, already presented such an enigma. Shapley presented
these dwarfs as a new type of \emph{stellar system} resembling the
Magellanic Clouds and globular clusters but did not attempt to speculate
on their exact nature. While dwarf galaxies were quickly understood to
be galaxies based on the inferred luminosities and sizes, their nature
and formation remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
(dSph) galaxies \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987},
inferred mass-to-light ratios were at least 10 times larger than the
values of globular clusters (GCs). While rather uncertain initially,
these values have been corroborated with larger and more precise samples
\citep[e.g.,][]{hargreaves+1994}. At the time, several theories were
proposed to explain these unusually high mass-to-light ratios. Examples
include: ongoing tidal disruption inflating inferred velocity
dispersions \citep[e.g.,][]{kuhn+miller1989}, the presence of massive
central black holes \citep[e.g.,][]{strobel+lake1994}, or modified
theories of gravity \citep{milgrom1995}. Over time, however, a consensus
developed where the high mass-to-light ratios of dwarf galaxies was due
to the presence of a dark matter halo, which was lacking in GCs
\citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}. Since then, the
properties of dwarf galaxies have played an increasingly important role
in our understanding of the clustering of dark matter on small scales
\citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.

Today, a common definition for a dwarf galaxy is a gravitationally bound
stellar system with dark matter \citep[or, more specifically, not
consistent with newtonian dynamics of visible matter
alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
have no clear evidence for dark matter. The boundary between these two
classes becomes uncertain for faint, compact systems, known as
``ambiguous'' systems.

The Local Group (LG) is defined as the group consisting of galaxies
within \(\sim 1\) Mpc from the Milky Way-Andromeda centre \citep[e.g.,][
and references therein]{mcconnachie2012}. Today, we know that the Milky
Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, globular clusters, and ambiguous systems. Such
a nearby population of dwarf galaxies are useful for resolved, detailed
studies aiming to understand the nature of these systems.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\) or stellar masses
\(10^9\,\Mo \gtrsim M_\star \gtrsim 3\times10^7\,\Mo\) \footnote{assuming
  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
  populations}, often exhibit irregular morphologies and recent star
formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
(LMC) as an example of an irregular, bright dwarf galaxy, where most
stars are in a rotationally-supported thin disk (seen nearly face-on)
with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
intermediate luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\) or
\(3\times10^7\,\Mo \gtrsim M_\star \gtrsim 10^5\,\Mo\)). Typically,
these systems are old, non-star forming, gas-poor, and spheroidal. All
Milky Way satellites discovered before digital sky surveys are
classicals. The 12 classical dwarf satellites of our Galaxy are (in
order of decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor,
Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans
I, and Crater II.\footnote{While formally the dwarf galaxy names we
  discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor dSph, we
  omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
occupy the very faintest magnitude range (\(-7.7 \lesssim M_V\), or
\(M_\star < 10^5\,\Mo\)). These galaxies have minuscule stellar masses,
typically compact sizes, and very metal-poor stellar populations
\citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
more than 15 orders in magnitude, or over 7 decades in stellar mass.

Here, we focus on classical dwarf galaxies, which are ideal for detailed
dynamical study. These galaxies are dark matter dominated but also have
large numbers of stars with radial velocities and proper motions, and
have generally well-constrained properties.

@@

\section{Cosmological context}\label{cosmological-context}

We only understand a fraction of the universe's composition. The leading
theory of cosmology, \LCDM{} (Lambda Cold Dark Matter), posits that the
universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
matter (DM), and 5\% regular baryons\footnote{\emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy drives the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
\citep[e.g.,][]{hayashi+2023}.

In \LCDM{}, dark matter is assumed to interact only gravitationally.
Light passes through dark matter unimpeded---in this sense dark matter
is transparent. Dark matter is also commonly assumed to be \emph{cold},
i.e.~typical velocities much smaller than the speed of light in the
early universe. If dark matter is cold, then it should condense on all
scales, forming non-linear structures (or \emph{halos}) from the size of
galaxy clusters to smaller than the faintest dwarf galaxies.
Implications of dark matter properties include cosmological structural
formation, galaxy structure, and galaxy interactions.

@@

\subsection{Typical structure of dark matter
halos}\label{typical-structure-of-dark-matter-halos}

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile has shown remarkable success in describing \LCDM{}
halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
radii \(r \ll r_s\). The steepness of the density profile increases
gradually with radius, and at large radii the density falls off like
\(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density} shows an example NFW
halo.

The total mass of an NFW profile formally diverges, so halo masses are
commonly defined using an overdensity criterion. The virial mass,
\(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
containing a mean enclosed density 200 times\footnote{For the collapse
  of a uniform spherical density, the virialized overdensity would be
  \(\Delta = 18\pi^2\approx 178\) for a critical universe
  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
  this parameter may be closer to \(\Delta \approx 100\) for our
  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
  from][]{bryan+norman1998}.} the critical density of the universe:
\begin{equation}{
M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G.
}\end{equation} A standard second parameter of NFW halos is the halo
concentration, \(c=r_{200} / r_s\), which describes how the
characteristic radius scale of the halo compares to the virial radius.
Using this parameter, the scale density is a function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}.

@@

The two parameters of an NFW profile are not independent. Lower-mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass \citep[e.g.,][]{bullock+2001, ludlow+2016}. The
left panel of Fig.~\ref{fig:smhm} illustrates the present-day
mass-concentration from \citet{ludlow+2016}. While concentration tends
to decrease with increasing mass, the relation has substantial scatter.
Other parameters such as the halo spin or shape may affect the scatter
of the mass-concentration relation, but their effect is typically
expected to be small
\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Sculptor-like galaxy.
The dark matter is more extended and massive than the star across the
entire galaxy.}\label{fig:nfw_density}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
\caption[Stellar-mass halo-mass relation]{\textbf{Left} The NFW halo
concentration \(c=r_{200} / r_s\) as a function of virial mass
\(M_{200}\). The solid line with 1-\(\sigma\) shaded region is the
mass-concentration relation from \citet{ludlow+2016} for \(z=0\).
\textbf{Right} Stellar mass (top) as a function of maximum circular
velocity. The solid line with the 1-\(\sigma\) shaded region is the
relation from \citet{fattahi+2018} with scatter points simulated central
galaxies from \apostle{} in \citet{fattahi+2018}.}\label{fig:smhm}
\end{figure}

\subsection{\texorpdfstring{Formation of galaxies within \LCDM{}
halos}{Formation of galaxies within  halos}}\label{formation-of-galaxies-within-halos}

@@

In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxies
comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
shows an example exponential stellar component for a classical dwarf
galaxy with the expected underlying dark matter halo. Even when the
stars are most dense in the inner regions, the dark matter remains
nearly an order of magnitude higher in density. Stars have a small
contribution to the gravitational structure of dwarf galaxies--instead
stars are reasonably approximated as tracer particles of the underlying
dark matter halo. In addition, the stellar component is much more
compact than the dark matter halo.

Several challenges complicate the SMHM trend, including environment,
assembly history, tidal effects, and the details of galaxy formation. In
particular, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from host's circumgalactic medium) and
tidal removal of gas cause star formation to quench
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for low-mass
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on Local Group dwarf galaxies is essential for determining where
and how these galaxies formed in a cosmological context.

\section{Observing faint features of dwarf
galaxies}\label{observing-faint-features-of-dwarf-galaxies}

\subsection{\texorpdfstring{The era of
\emph{Gaia}}{The era of Gaia}}\label{the-era-of-gaia}

Since Local Group dwarfs are nearby, they are resolved into individual
stars, and we can study these galaxies on a star-by-star basis. As a
result, it is possible to measure the 3D velocity and position of a star
if we can measure the its position, distance, line-of-sight (LOS)
velocity, and proper motions. Unfortunately, determining distances and
full 3D velocities is challenging. The most direct measurement of
distance, parallax, requires precise tracking of a star's sky position
across a year. And while line-of-sight (LOS) velocities are relatively
easily determined from spectroscopy, tangential velocities, derived from
proper motions and distances, are much more challenging. Typically,
measuring proper motions requires accurate (much less than arcsecond)
determinations of small changes in a star's position over baselines of
years to decades. The full 6D position and velocity information for
stars has, until recently, been known for only a handful of stars.

Launched in 2013, \emph{Gaia} is a space-based, all-sky survey telescope
situated at the Sun-Earth L2 Lagrange point
\citep{gaiacollaboration+2016}. \emph{Gaia} has redefined astrometry,
providing photometry, positions, proper motions, and parallaxes for over
1 billion stars \citep{gaiacollaboration+2021}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
are expected.

Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived. In addition to astrometry,
\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
and colours from the blue photometer (BP, 330-680 nm) and red photometer
(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
BP-RP spectra and radial velocity measurements of bright stars \citep[of
magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension, \(\alpha\), and declination, \(\delta\),
  typically in units of mili-arcsecond (mas) per year.
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

\subsection{The Local Group of
Galaxies}\label{the-local-group-of-galaxies}

\emph{Gaia} has revolutionized LG studies. For example, the 6D dynamical
measurements and metallicities of MW stars led to the (re)discovery of
past mergers or Milky Way building blocks like \emph{Gaia}-Sausage
Enceladus
\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
see also][]{meza+2005}}, out-of-equilibrium structures like the
\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
Way halo, \emph{Gaia} has helped find and constrain numerous stellar
streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
Altogether, \emph{Gaia} has revealed the hierarchical formation and
complex, evolving structure of our own Galaxy.

For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
facilitated stellar membership determination. Before \emph{Gaia}, few
galaxies had precisely measured proper motions, more recently from
Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
\emph{Gaia} allowed the first systematic and precise determinations of
Milky Way satellite proper motions \citep{pace+li2019, MV2020a}. While
the proper motion uncertainty of a typical dwarf member star is often
large, by combining the proper motions of 100s or 1000s of stars from
\emph{Gaia}, precise average proper motion measurements can be
determined, sometimes only limited by \emph{Gaia}'s systematic error
floor \citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a
new dynamical era for MW satellite studies, where we can derive precise
orbits for any satellite, assuming a given MW potential. In addition,
\emph{Gaia} helps establish membership by separate out contaminating MW
foreground stars. By measuring parallaxes and/or proper motions, many
background and foreground stars can be classified as non-members
\citep[e.g.,][]{battaglia+2022, jensen+2024}.

\subsection{Dwarf galaxies today}\label{dwarf-galaxies-today}

Advances in observational techniques has accelerated progress in our
understanding of dwarf galaxies and introduced new questions. Deep
digital photometric surveys have more than quadrupled the number of
known Milky Way dwarf galaxies since 2005 \citep{simon2019}. Upcoming
surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
Time \citep[LSST,][]{ivezic+2019}, will likely identify even fainter
dwarf galaxies. In addition, large aperture multi-object spectrographs
have revealed the detailed and complex inner chemodynamical structure of
dwarf galaxies \citep[e.g.,][]{battaglia+2006, fabrizio+2016}. Beyond
precise structural and kinematic properties of dwarf galaxies, modern
observations allow for the separation of multiple stellar populations,
detailed constraints on the dark matter density profiles, and hints of
tidal disruption or stellar halos.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These galaxies are typically defined to be
fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
or similarly \(M_\star \lesssim 10^9 M_\odot\)
\citep[e.g.,][]{hodge1971, mcconnachie2012}. Because the galaxy
luminosity function increases towards fainter objects, dwarfs are the
most numerous galaxies in the Universe
\citep[e.g.,][]{blanton+2005, mao+2021}. Dwarf galaxies are also highly
\emph{dark-matter dominated}, with mass to light ratios which may exceed
1000 \(M_\odot/ L_\odot\) \citep[e.g.,][]{simon+geha2007, hayashi+2023}.

With the exception of the Magellanic Clouds, most dwarf galaxy
satellites of the Milky Way (MW) are \emph{quenched}, with little to no
recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
satellites contain old stellar populations which are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars known. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
small scales to the formation of metal-poor stellar populations.

In this Chapter, we first describe the general observed properties of
local dwarf galaxies. Next, we summarize our understanding of the
cosmological origin of dwarf galaxies. We later review recent
advancements and pending questions concerning dwarf galaxies, and
introduce the puzzle posed by the extended stellar density profiles of
Sculptor and Ursa Minor. We end with a brief roadmap to the remainder of
this dissertation.

\section{Observations of dwarf
galaxies}\label{observations-of-dwarf-galaxies}

Dwarf galaxies have long raised conundrums for theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) are also classified as dwarf galaxies, but these
  were likely always known to humans at southern latitudes.}, with no
known analogues, already presented such an enigma. Shapley presented
these dwarfs as a new type of \emph{stellar system} resembling the
Magellanic Clouds and globular clusters but did not attempt to speculate
on their exact nature. While dwarf galaxies were quickly understood to
be galaxies based on the inferred luminosities and sizes, their nature
remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor (Scl) and Ursa Minor (UMi) dwarf
spheroidal (dSph) galaxies, inferred mass-to-light ratios were at least
10 times larger than the values of globular clusters {[}GCs; e.g.,
\citet{aaronson1983}; \citet{aaronson+olszewski1987}{]}. While rather
uncertain initially, these values were later corroborated with larger
and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the time,
several theories were proposed to explain these unusually high
mass-to-light ratios. Examples include: ongoing tidal disruption
inflating inferred velocity dispersions
\citep[e.g.,][]{kuhn+miller1989}, the presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
\citep{milgrom1995}. Over time, however, a consensus developed where the
high mass-to-light ratios of dwarf galaxies was due to the presence of a
dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
Since then, the properties of dwarf galaxies have played an increasingly
important role in our understanding of the clustering of dark matter on
small scales \citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.

Today, a common definition for a dwarf galaxy is a gravitationally bound
stellar system with dark matter \citep[or, more specifically, not
consistent with Newtonian dynamics of visible matter
alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
have no clear evidence for dark matter. The boundary between these two
classes becomes uncertain for faint, compact stellar associations, known
as ``ambiguous'' systems.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes \(-14 \gtrsim M_V \gtrsim  -18\)
or stellar masses
\(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
  populations}, often exhibit irregular morphologies and recent star
formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
(LMC) as an example of an irregular, bright dwarf galaxy, where most
stars are in a rotationally-supported thin disk (seen nearly face-on)
with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
intermediate luminosities ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
\(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
these systems are old, non-star forming, gas-poor, and spheroidal. All
Milky Way satellites discovered before digital sky surveys are
classicals. The 12 classical dwarf satellites of our Galaxy are (in
order of decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor,
Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans
I, and Crater II.\footnote{While formally the dwarf galaxy names we
  discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor dSph, we
  omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
occupy the very faintest magnitude range (\(M_V \gtrsim -7.7\) or
\(M_\star \lesssim 10^5\,\Mo\)). These galaxies have minuscule stellar
masses, typically compact sizes, and very metal-poor stellar populations
\citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
more than 15 orders in magnitude, or over 7 decades in stellar mass.

Of local dwarfs, the classical systems still remain among the most
studied galaxies with the best constrained parameters. These dwarfs also
contain large numbers of bright (giant) stars which can be followed up
spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
the determination of fundamental properties such as the position, size,
orientation, distances, proper motions, line-of-sight (LOS) velocities,
and velocity dispersions \(\sigma_\V\), are all relatively well
constrained today. \textbf{revisit this\ldots JFN cuts a similar
version}

Most well-studied dwarf galaxies lie near the Milky Way or in its
vicinity, the Local Group (LG) of galaxies. TheLG is defined as the
group consisting of galaxies within \(\sim 1\) Mpc from the Milky
Way-Andromeda centre \citep[e.g.,][ and references
therein]{mcconnachie2012}. Today, we know that the Milky Way system is
teeming with dwarfs, many of which are satellites of either the Milky
Way or(MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
the MW satellite system, including dwarf galaxies, globular clusters,
and ambiguous systems. Such a nearby population of dwarf galaxies are
useful for resolved, detailed studies aiming to understand the nature of
these systems.

@@

\section{Dwarf galaxies in a cosmological
context}\label{dwarf-galaxies-in-a-cosmological-context}

We only understand a fraction of the universe's composition. The leading
theory of cosmology, Lambda Cold Dark Matter (\LCDM{}), posits that the
universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
matter (DM), and 5\% regular baryons\footnote{\emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy drives the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
\citep[e.g.,][]{hayashi+2023}.

In \LCDM{}, dark matter is assumed to interact only gravitationally.
Light passes through dark matter unimpeded---in this sense, dark matter
is transparent. Dark matter is also commonly assumed to be \emph{cold},
i.e.~typical velocities much smaller than the speed of light in the
early universe. If dark matter is cold, then it should condense on all
scales, forming non-linear structures (or \emph{halos}) from the size of
galaxy clusters to smaller than the faintest dwarf galaxies.
Implications of dark matter properties include cosmological structural
formation, galaxy structure, and galaxy interactions.

@@

\subsection{The structure of cold dark matter
halos}\label{the-structure-of-cold-dark-matter-halos}

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile has shown remarkable success in describing \LCDM{}
halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
radii \(r \ll r_s\). The steepness of the density profile increases
gradually with radius, and at large radii the density falls off like
\(\rho \sim 1/r^3\).

The total mass of an NFW profile formally diverges, so halo masses are
conventionally defined using an overdensity criterion. The virial mass,
\(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
containing a mean enclosed density 200 times\footnote{For the collapse
  of a uniform spherical density, the virialized overdensity would be
  \(\Delta = 18\pi^2\approx 178\) for a critical universe
  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
  this parameter may be closer to \(\Delta \approx 100\) for our
  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
  from][]{bryan+norman1998}.} the critical density of the universe:
\begin{equation}{
M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G,
}\end{equation} and \(H(z)\) is the Hubble constant as a function of
redshift. Another way of characterizing NFW halos is through the halo
concentration, \(c=r_{200} / r_s\), which describes how the
characteristic radius scale of the halo compares to the virial radius.
Using this parameter, the scale density is a function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}.

@@

The blue solid curve in Fig.~\ref{fig:nfw_density} shows an example NFW
halo for a galaxy with \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
\(M_{200} = 1\times10^10\,\Mo\) and \(c=12.5\), the cosmological mean
for a Fornax stellar mass galaxy.

The two parameters of an NFW profile are not independent. Lower-mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass or, equivalently, the dependence of \(\vmax\) on
\(\rmax\) \citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration from
\citet{ludlow+2016}. While concentration tends to decrease with
increasing mass, the relation has substantial scatter. Other parameters
such as the halo spin or shape may affect the scatter of the
mass-concentration relation, but their effect is typically expected to
be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Fornax-like galaxy. The
dark matter is more extended and massive than the star across the entire
galaxy.}\label{fig:nfw_density}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
\caption[Stellar-mass halo-mass relation]{\textbf{Left} The NFW halo
concentration \(c=r_{200} / r_s\) as a function of virial mass
\(M_{200}\). The solid line with 1-\(\sigma\) shaded region is the
mass-concentration relation from \citet{ludlow+2016} for \(z=0\).
\textbf{Right} Stellar mass (top) as a function of maximum circular
velocity. The solid line with the 1-\(\sigma\) shaded region is the
relation from \citet{fattahi+2018} with scatter points simulated central
galaxies from \apostle{} in \citet{fattahi+2018}. The pink star
illustrates the location of the Fornax galaxy, whose density profiles
are shown in Fig.~\ref{fig:nfw_density}.}\label{fig:smhm}
\end{figure}

\subsection{\texorpdfstring{Galaxies formation in
\LCDM{}}{Galaxies formation in }}\label{galaxies-formation-in}

@@

In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxy
comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
shows an example exponential stellar component for the Fornax dwarf
galaxy with the expected underlying dark matter halo. This galaxy has a
stellar mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light
radius \(0.65\,\kpc\) \citep[ with \citet{woo+courteau+dekel2008}
mass-to-light ratio]{munoz+2018}, and a dark matter halo with
\(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\). Even when the stars are
densest (the inner regions), the dark matter remains nearly an order of
magnitude higher in density. Stars make a small contribution to the
gravitational structure of dwarf galaxies--indeed, stars are reasonably
approximated as tracer particles of the underlying dark matter halo. In
addition, the stellar component is confined to the central regions of
the dark matter halo.

Several factors affect the SMHM trend, including environment, assembly
history, tidal effects, and the details of galaxy formation. In
particular, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from the host's circumgalactic medium) and
tidal removal of gas cause star formation to quench
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for low-mass
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on Local Group dwarf galaxies is essential for determining where
and how these galaxies formed in a cosmological context.

\subsection{Open questions and ongoing challenges in dwarf
galaxies}\label{open-questions-and-ongoing-challenges-in-dwarf-galaxies}

Advances in observational techniques have accelerated progress in our
understanding of dwarf galaxies and introduced new questions. Deep
digital photometric surveys have more than quadrupled the number of
known Milky Way dwarf galaxies since 2005 \citep{simon2019}. Upcoming
surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
Time \citep[LSST,][]{ivezic+2019}, will likely identify even fainter
dwarf galaxies. In addition, large aperture multi-object spectrographs
have revealed the detailed and complex inner chemodynamical structure of
dwarf galaxies \citep[e.g.,][]{battaglia+2006, fabrizio+2016}. Beyond
precise structural and kinematic properties of dwarf galaxies, modern
observations allow for the separation of multiple stellar populations,
detailed constraints on the dark matter density profiles, and hints of
tidal disruption or stellar halos. \textbf{does this ¶ (or section)
belong here?}
